Integrating brain and biomechanical models - a new paradigm for understanding neuro-muscular control

### 2017-11-18 papers/integration/omm_integ.tex
Further WIP

There is a long history of modelling the oculomotor system. For a
comprehensive review, see \cite{girard_brainstem_2005}. Models of
individual \rvw{subsystems} have been proposed for brainstem
\citep{robinson_oculomotor_1975,scudder_new_1988,gancarz_neural_1998},
cerebellum \citep{quaia_model_1999,dean_modelling_1995,dean_learning_1994}
and superior colliculus
\citep{arai_two-dimensional_1994,moren_mechanism_2013,marino_spatial_2012}.
More recently, combined models have also been developed incorporating
sensory input \citep{cope_basal_2017} and driving \rvw{a second order
differential equation} representing the eye
\citep{tabareau_geometry_2007,nguyen_saccade_2014,thurat_biomimetic_2015}.
None of these models has yet fully closed the loop to produce a
behaving system operating freely within its environment. We argue that
developing integrated, closed-loop models of behaving systems offers
insights into the operation of neural systems that are not available
from input-~or output-assumption models.

@@

To the \ccg~model, we added a brainstem model and a biomechanical eye
model. \rvw{The rate-coded brainstem model was taken from the
literature} \citep{gancarz_neural_1998} as the best-of-breed saccadic
burst generator \citep{girard_brainstem_2005}.  The biomechanical eye
was implemented using the biomechanical modelling framework OpenSim.
These will be described below, \rvwt{along with a review of
the \ccg~model}, but first we will give a description of the
co-ordinate systems and the modelling framework used.

\subsubsection{Co-ordinates in the world} \label{sec:methods:coords}

@@

\rvwt{One hypothesis for the r\^ole played by the basal ganglia (BG)
is that the system performs \e{action selection}
\citep{mink_basal_1996,redgrave_basal_1999,hikosaka_role_2000}.
The \ccg~model places the BG at the centre of the oculomotor system,
following the known anatomy of the
region \citep{hikosaka_role_2000}. The BG receives input indirectly
from the superior colliculus, which has a retinotopic
arrangement \citep{ottes_visuomotor_1986}.
%and a signal which is associated with saccadic eye
%movements \citep{wurtz_superior_1971}, specifying the saccade
%end-point.
The BG receives excitatory
inputs directly from retinotopic regions of the cortex including the
frontal eye fields (FEF), supplementary eye fields (SEF), lateral
intraparietal cortex (LIP) and dorsolateral prefrontal cortex. The
dorsolateral prefrontal cortex, which participates in voluntary
saccades
\citep{funahashi_prefrontal_1993,munoz_look_2004}, is not
modelled because the model concerns reflexive rather than voluntary
eye movements. Several other regions of the brain that are associated
with eye movements are also omitted from the model. The early visual
processing stream in cortex, from V1, through to the LIP is subsumed
into a `sustained retinal' signal. The justification here is that the
model reacts to simple luminant targets and does not need to carry out
the feature extraction performed by these visual areas. The
supplementary eye fields are involved in the programming of saccade
sequences \citep{tehovnik_eye_2000} and memory guided
saccades \citep{chen_supplementary_1995,schlag_neurons_2002}. Lesions
of SEF do not affect visually guided
saccades \citep{gaymard_cortical_1998} and so the SEF is also omitted
from the model.}
%
% Figure here is the ``macroscopic architecture'' one.
%

% Now describe each bit a little more, with help of the figure.
\rvwt{Fig.~\ref{fig:ccg_model}(a) shows the macroscopic architecture of
the \ccg~model. The figure shows the relationships between the retinal
input populations, the FEF, the populations comprising the BG
sub-system (the red border indicates that the box represents a number
of populations as a sub-system), the thalamus and the superior
colliculus. Excitatory connections are indicated with arrow heads;
inhibitory connections with circles in place of the arrow heads. The
blue and green connection lines indicate two thalamo-basal ganglia
loops, one cortical loop through FEF (green), the other a sub-cortical
loop through SC. It is important to note that although they are given
different colours in the diagram, these loops are in no way
independent, with loop activity combining both in thalamus and in the
basal ganglia and a direct excitatory, feed-forward connection from
FEF to SC.}

% More on BG model here
\rvwt{The basal ganglia sub-system is the most complex component of
the \ccg~model. The BG model is based on previous
work \citep{gurney_computational_2001,gurney_computational_2001-1} and is referred to as the GPR
model. The GPR model incorporates the following main components of the
primate BG \citep{mink_basal_1996,wickens_basal_1997}: (i) The
striatum (the main input station to the BG) which is divided into two
iterdigitated poulations of projection neurons expressing primarily D1
or D2-type dopaminergic receptors (named Str\_D1 and Str\_D2); (ii)
The subthalamic nucleus (STN); (iii) the external segment of the
globus pallidus (GPe); (iv) the output nucleus relevant for saccadic
control---the substantia nigra pars reticulata
(SNr) \citep{hikosaka_role_2000}.}

\rvwt{The connectivity of the GPR model (Fig.~\ref{fig:ccg_model}(b)
is constrained by the known anatomy and physiology of the
BG~\citep{bolam_synaptic_2000}. Physiologically, the only source of
glutamate within the BG is the STN, whose projections are therefore
excitatory; all other nuclei have GABAergic projection neurons and are
therefore inhibitory. The cortex sends glutamatergic projections to
both the Str\_D1 striatal population, which projects preferentially to
the SNr, and to Str\_D2, which projects primarily to
GPe \citep{gerfen_d1_1990}. The cortex also projects to the STN, which
sends diffuse projections to the SNr and
GPe \citep{parent_anatomical_1993}. The GPe projects to the SNr and
also projects back to the STN, completing a GPe--STN loop.}

\rvwt{Within the BG, there are several mechanisms supporting
competitive processing for selecting channels whose inhibitory output
is reduced. The selection mechanism of the GPR model is the
`off-centre, on-surround' scheme proposed
by \cite{mink_basal_1993}. The `on-surround' is provided by
diffuse, excitatory projections from the STN to the SNr. Focussed
inhibition from the Str\_D1 neurons in striatum contributes the
`off-centre' part of the mechanism. This arrangement leads to
selection behaviour via a release of target inhibition, since
channels that have strong saliencel (input) have weak output at the
level of SNr, and channels with weak salience have enhanced output.}

\rvwt{The GPe is not included in the centre-surround circuit described
above, but still plays a key r\^ole in selection. Operating alone, the
Str\_D1/STN/SNr circuit can suffer from the following problem: if the
input for all channels is relatively high, then the diffuse projection
from STN, which effectively supplies a sum of the STN inputs to each
channel in SNr, will provide so much excitation that Str\_D1 may
become unable to inhibit one of the channels in SNr and selection may
become
impossible. \cite{gurney_computational_2001,gurney_computational_2001-1}
showed that the inhibitory feedback from GPe to STN acts as an
`automatic gain control' to help prevent this from occurring.}

\rvwt{At the neuronal level, the STN, GPe and SNr have tonic output
levels \citep{chevalier_disinhibition_1990,delong_primate_1985,kita_intracellular_1991}. This
is modelled using piecewise linear output functions with zero offsets,
$c$ (see Eq.~\ref{eq:lintransfunc}). In striatum, Str\_D1 and Str\_D2
have positive offset $c$, mimicking the so-called `down-state' of
medium spiny neurons which have a resting potential far below spiking
threshold and require co-ordinated input to generate action
potentials \citep{wilson_origins_1996}.
% Note, in Cope's paper, there is an error with the offsets for
% D1,D2. These are set to -0.05 instead of 0.05. A quick trial with
% the model suggests this doesn't make a very large difference to the
% operation, and I have the sign correct in any case.
In addition, the Str\_D1 and Str\_D2 neurons are influenced by
dopamine in different ways; facilitating cortico-striatal transmission
at medium spiny neurons with D1 receptors \citep{hernandez-lopez_D1_1997,gonon_prolonged_1997} and reducing
transmission at those with D2 receptors \citep{delgado_inhibitory_1999}. These effects are
modelled using a dopamine parameter which modulates the input
activation as described below in Eqs~\ref{eq:a_in_d1} \& \ref{eq:a_in_d2}.}

% Change this - we already talked about channels?
\rvwt{The \ccg~model adopts a `channel based' model from previous
work~\citep{redgrave_basal_1999,gurney_computational_2001}.
Within the context of the basal ganglia model, a `channel' is a
sub-set of neurons which are associated with an individual
action. Each channel in BG loops back via relays in the thalamus
through cortex. Input into this thalamo-cortical loop is termed
`salience'. Because it is a loop, input salience to the channel may
originate cortically or sub-cortically. The higher the salience of the
channel, the greater is the behavioural relevance of the associated
action.}

\rvwt{In the oculomotor model, an action
corresponds to a distinct saccade end point. It can be seen that the
action space for reflexive saccade end points is therefore
continuous rather than discrete. MORE HERE}

% FEF stuff taken from Alex's pre-pub

Fig.~\ref{brain_model} shows the layout of the populations in the
brain model, as implemented in SpineCreator, and the interconnections
between them. Each population of 2500 elements is arranged as a 50 by
50 grid which forms a map in a retinotopic co-ordinate system, roughly
matching the known layout of the superior
colliculus \citep{robinson_eye_1972}.
% and frontal eye field?

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[width=0.9\textwidth]{./figures/ccg_model.png}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:ccg_model} Figure \arabic{figure}.}
{ The macroscopic architecture of the \ccg~model. The main nuclei
modelled as brain systems are: basal ganglia (BG), frontal eye fields
(FEF), thalamus (THAL) and superior colliculus (SC). The retinal input
is presented via non-biologically plausible retinal populations. The
loops through basal ganglia, which define the architecture, are shown
with coloured lines: the cortical loop (through FEF and THAL) in green
and the sub-cortical loop (through SC and THAL) in blue. Connections
with arrowheads indicate excitatory connections, those with circles
are inhibitory. A red border indicates that the box represents a
sub-system of two or more populations; a black box indicates,
within the context of the model, a single neural population.}
\end{figure}

>>>>

There is a long history of modelling the oculomotor system. For a
comprehensive review, see \cite{girard_brainstem_2005}. Models of
individual \rvw{sub-systems} have been proposed for brainstem
\citep{robinson_oculomotor_1975,scudder_new_1988,gancarz_neural_1998},
cerebellum \citep{quaia_model_1999,dean_modelling_1995,dean_learning_1994}
and superior colliculus
\citep{arai_two-dimensional_1994,moren_mechanism_2013,marino_spatial_2012}.
More recently, combined models have also been developed incorporating
sensory input \citep{cope_basal_2017} and driving \rvw{a second order
differential equation} representing the eye
\citep{tabareau_geometry_2007,nguyen_saccade_2014,thurat_biomimetic_2015}.
None of these models has yet fully closed the loop to produce a
behaving system operating freely within its environment. We argue that
developing integrated, closed-loop models of behaving systems offers
insights into the operation of neural systems that are not available
from input-~or output-assumption models.

@@

To the \ccg~model, we added a brainstem model and a biomechanical eye
model. \rvw{The rate-coded brainstem model was taken from the
literature} \citep{gancarz_neural_1998} as the best-of-breed saccadic
burst generator \citep{girard_brainstem_2005}.  The biomechanical eye
was implemented using the biomechanical modelling framework OpenSim;
the brain and brainstem were modelled using the SpineML toolchain.
These will be described below, \rvwt{along with a review of
the \ccg~model}, but first we will give a description of the
co-ordinate systems.

\subsection{Co-ordinates in the world} \label{sec:methods:coords}

@@

\rvwt{One hypothesis for the r\^ole played by the basal ganglia (BG)
is that the system performs \e{action selection}
\citep{mink_basal_1996,redgrave_basal_1999,hikosaka_role_2000}.
The \ccg~model places the BG at the centre of the oculomotor system,
following the known anatomy of the
region \citep{hikosaka_role_2000}. The BG receives input indirectly
from the superior colliculus, which has a retinotopic
arrangement \citep{ottes_visuomotor_1986}.
%and a signal which is associated with saccadic eye
%movements \citep{wurtz_superior_1971}, specifying the saccade
%end-point.
The BG receives excitatory
inputs directly from retinotopic regions of the cortex including the
frontal eye fields (FEF), supplementary eye fields (SEF), lateral
intraparietal cortex (LIP) and dorsolateral prefrontal cortex. The
dorsolateral prefrontal cortex, which participates in voluntary
saccades
\citep{funahashi_prefrontal_1993,munoz_look_2004}, is not
modelled because the model concerns reflexive rather than voluntary
eye movements. Several other regions of the brain that are associated
with eye movements are also omitted from the model. The early visual
processing stream in cortex, from V1, through to the LIP is subsumed
into a `sustained retinal' signal which arrives at FEF. The
justification here is that the model reacts to simple luminant targets
and does not need to carry out the feature extraction performed by
these visual areas. The supplementary eye fields are involved in the
programming of saccade sequences \citep{tehovnik_eye_2000} and memory
guided
saccades \citep{chen_supplementary_1995,schlag_neurons_2002}. Lesions
of SEF do not affect visually guided
saccades \citep{gaymard_cortical_1998} and so the SEF is also omitted
from the model.}

% Now describe each bit a little more, with help of the CCG
% macroscopic architecture figure.
\rvwt{Fig.~\ref{fig:ccg_model}(a) shows the macroscopic architecture of
the \ccg~model. The figure shows the relationships between the retinal
input populations, the FEF, the populations comprising the BG
sub-system (the red border indicates that the box represents a number
of populations as a sub-system), the thalamus and the superior
colliculus. Excitatory connections are indicated with arrow heads;
inhibitory connections with circles in place of the arrow heads. The
blue and green connection lines indicate two thalamo-basal ganglia
loops, one cortical loop through FEF (green), the other a sub-cortical
loop through SC. It is important to note that although they are given
different colours in the diagram, these loops are in no way
independent, with loop activity combining both in thalamus and in the
basal ganglia and a direct excitatory, feed-forward connection from
FEF to SC.}

% More on BG model here
\rvwt{The basal ganglia sub-system is the most complex component of
the \ccg~model. The BG model is based on previous
work \citep{gurney_computational_2001,gurney_computational_2001-1} and is referred to as the GPR
model. The GPR model incorporates the following main components of the
primate BG \citep{mink_basal_1996,wickens_basal_1997}: (i) The
striatum (the main input station to the BG) which is divided into two
iterdigitated populations of projection neurons expressing primarily D1
or D2-type dopaminergic receptors (named Str\_D1 and Str\_D2); (ii)
The subthalamic nucleus (STN); (iii) the external segment of the
globus pallidus (GPe); (iv) the output nucleus relevant for saccadic
control---the substantia nigra pars reticulata
(SNr) \citep{hikosaka_role_2000}.}

\rvwt{The connectivity of the GPR model [Fig.~\ref{fig:ccg_model}(b)]
is constrained by the known anatomy and physiology of the
BG~\citep{bolam_synaptic_2000}. Physiologically, the only source of
glutamate within the BG is the STN, whose projections are therefore
excitatory; all other nuclei have GABAergic projection neurons and are
therefore inhibitory. The cortex sends glutamatergic projections to
both the Str\_D1 striatal population, which projects preferentially to
the SNr, and to Str\_D2, which projects primarily to
GPe \citep{gerfen_d1_1990}. The cortex also projects to the STN, which
sends diffuse projections to the SNr and
GPe \citep{parent_anatomical_1993}. The GPe projects to the SNr and
also projects back to the STN, completing a GPe--STN loop.}

\rvwt{The GPR model is arranged into `action channels';
Fig.~\ref{fig:ccg_model}(b) shows an example network containing three
channels. It is between these channels that competition occurs, with
the winning channel succeeding in reducing activity in the output
nucleus, SNr, and thereby disinhibiting its target. The complete
connectivity pattern for this small network is shown in
Fig.~\ref{fig:ccg_model}(b); the left channel in cortex innervates the
left channels of Str\_D1, STN and Str\_D2. Connections are one-to-one,
so it follows that the middle channel of cortex innervates the middle
channels of STN and the striatal populations and the right channel of
cortex innervates right channels in striatum and STN. Striatal
population channels also inhibit SNr and GPe on a one-to-one basis and
GPe feeds inhibition to SNr and STN in a one-to-one manner. The
outputs from STN however are not one-to-one. The output from all
channels of STN is summed together and then the sum is fed into each
channel of SNr and GPe. This models the diffuse excitation from STN
which has been observed in the BG \citep{parent_anatomical_1993}.}

\rvwt{Within the BG, there are several mechanisms supporting
competitive processing for selecting channels whose inhibitory output
should be reduced. The selection mechanism of the GPR model is the
`off-centre, on-surround' scheme proposed
by \cite{mink_basal_1993}. The `on-surround' is provided by
diffuse, excitatory projections from the STN to the SNr. Focussed
inhibition from the Str\_D1 neurons in striatum contributes the
`off-centre' part of the mechanism. This arrangement leads to
selection behaviour via a release of target inhibition, since
channels that have strong salience (input) have weak output at the
level of SNr, and channels with weak salience have enhanced output.}

\rvwt{The GPe is not included in the centre-surround circuit described
above, but still plays a key r\^ole in selection. Operating alone, the
Str\_D1/STN/SNr circuit can suffer from the following problem: if the
input for all channels is relatively high, then the diffuse projection
from STN, which effectively supplies a sum of \e{all} of the STN inputs to each
channel in SNr, will provide so much excitation that Str\_D1 may
become unable to inhibit one of the channels in SNr and selection may
become
impossible. \cite{gurney_computational_2001,gurney_computational_2001-1}
showed that the inhibitory feedback from GPe to STN acts as an
`automatic gain control' to help prevent this from occurring.}

\rvwt{At the neuronal level, the STN, GPe and SNr have tonic output
levels \citep{chevalier_disinhibition_1990,delong_primate_1985,kita_intracellular_1991}. This
is modelled using piecewise linear output functions with zero offsets,
$c$ (see Eq.~\ref{eq:lintransfunc}) but with noise added to the
input. In striatum, Str\_D1 and Str\_D2 have positive offset $c$,
mimicking the so-called `down-state' of medium spiny neurons which
have a resting potential far below spiking threshold and require
co-ordinated input to generate action
potentials \citep{wilson_origins_1996}.
% Note, in Alex's paper, there is an error with the offsets for D1,
% D2. These are set to -0.05 instead of 0.05. A quick trial with the
% model suggests this doesn't make a very large difference to the
% overall operation (you just see a background of noise around the
% signals in Str), and I have the sign correct in any case.
In addition, the Str\_D1 and Str\_D2 neurons are influenced by
dopamine in different ways; facilitating cortico-striatal transmission
at medium spiny neurons with D1 receptors \citep{hernandez-lopez_D1_1997,gonon_prolonged_1997} and reducing
transmission at those with D2 receptors \citep{delgado_inhibitory_1999}. These effects are
modelled using a dopamine parameter which modulates the input
activation as described below in
Eqs~\ref{eq:a_in_d1} \& \ref{eq:a_in_d2}; Str\_D1 activation is enhanced;
Str\_D2 activation is suppressed.}

% Change this - we already talked about channels in the context of the
% three channel model show. Perhaps modify to talk about how we have a
% map of channels - almost a continuous space in the \ccg model. This
% can then lead on to Fig 3.
\rvwt{Expand along lines of `the
action space for reflexive saccade end points is therefore
continuous rather than discrete' (FIXME).}

% Now some FEF stuff taken from Alex's pre-pub
\rvwt{Add description of FEF. (FIXME).}

Fig.~\ref{brain_model} shows the layout of the populations in the
brain model, as implemented in SpineCreator, and the interconnections
between them. \rvwt{Describe this figure in more detail. (FIXME).}
% and frontal eye field?

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[width=0.8\textwidth]{./figures/ccg_model.png}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:ccg_model} Figure \arabic{figure}.}
{ \bol{a) The macroscopic architecture of the \ccg~model.} The main
nuclei modelled as brain systems are: basal ganglia (BG), frontal eye
fields (FEF), thalamus (THAL) and superior colliculus (SC). The
retinal input is presented via non-biologically plausible retinal
populations. The loops through basal ganglia, which define the
architecture, are shown with coloured lines: the cortical loop
(through FEF and THAL) in green and the sub-cortical loop (through SC
and THAL) in blue. Connections with arrowheads indicate excitatory
connections, those with circles are inhibitory. A red border indicates
that the box represents a sub-system of two or more populations; a
black box indicates (at least, within the context of the model) a
single neural population. The BG box is expanded in: \bol{b) The basal
ganglia model component.}
%
This shows a basal ganglia comprising striatum (Str\_D1 \& Str\_D2),
subthalamic nucleus (STN), globus pallidum externum (GPe) and
substantia nigra reticulata (SNr). The model has three action channels
shown as black boxes within each blue population border. Three
channels of cortical input to the BG are also depicted. Red indicates
the activation level of a given channel, helping to illustrate the
selection mechanism. For example, the channel indicated by the
leftmost bar has a high salience (cortical input) and excites activity
in Str\_D1 which then inhibits the leftmost bar in SNr. The diffuse
projection from STN is equivalent to summing its projections
channel-wise, and then projecting the sum to all channels of its
target populations (the blue arrows indicate that all channels of GPe
and SNr are targeted by the connection). Dopaminergic modulation of
the inputs to the striatum are indicated by the blue circles labelled
`d' and the dotted lines. The SNr sends inhibitory output projections
to its targets.}
\end{figure}
